Polarized Cubical Types

### 2014-05-21 ring.tex
removed syntactic difference between 0 dim and higher dim

\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\denc}[1]{\llbracket #1 \rrbracket^c}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

%% double-blind reviewing...
\title{Cubical Polarized Types}
\authorinfo{}{}{}
\maketitle

@@

We use $t$ for regular first-order types which include the empty type, the
unit type, and conventional sum and product types. We call these types
$0$-dimensional. We use $\tau$ for cubical types of dimension 1 or higher:
these have the same syntactic structure as regular types except that they
also include \emph{negative} types. The syntax of the types is summarized
below:
\[\begin{array}{lrcl}
\textit{types of dimension $0$} & 
  t &::=& 0 \alt 1 \alt t_1 + t_2 \alt t_1 * t_2 \\
\textit{types of dimensions $\geq 1$} & 
  \tau &::=& \Box t \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The $\Box$ is used to make explicit the injection of a first-order type into
a 1-dimensional type. We omit the explicit injection when
convenient. Formally, the \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{t} &=& 0 \\
\bdim{\Box t} &=& 1 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \bdim{\tau}
\end{array}\]

It is easy to find a model for $0$-dimensional types: finite sets are such a
model. In more detail:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \star \} \\
\den{t_1 + t_2} &=& \den{t_1} \uplus \den{t_2} \\
\den{t_1 * t_2} &=& \den{t_1} \times \den{t_2} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of types maps to disjoint union of sets. The product of types maps to the
cartesian product of sets. We usually use $S$ for the denotation of these
conventional first-order types. Note that our (simple) semantics of types
identifies several structurally different types such as $(1+(1+1))$ and
$((1+1)+1)$. In some sense, this is innocent as the types are
isomorphic. However, in the operational semantics discussed in the next
section, we make the computational content of such type isomorphisms
explicit.

For cubical types of dimension $n \geq 1$, their denotations, usually written
as $\cubt$, are $n$-dimensional ``polarized cubes.'' Such an $n$-dimensional
cube $\nodet{\cubt_1}{\cubt_2}$ consists of a positive subspace $\cubt_1$ and
a negative subspace $\cubt_2$ (shaded in gray). Each of the subspaces are
cubes of a lower dimension; the $0$-dimensional case reduces to the
denotation of conventional first-order types. As we explain below, cubical
sets are closed under sums, differences, products.

A 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which intuitively
corresponds to the difference $t_1 - t_2$ (formally $\Box t_1 + (- \Box
t_2)$) of the two types whose denotations are $S_1$ and $S_2$ respectively. A
2-dimensional cube looks like
$\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

We now give the full denotation of cubical types:
\[\begin{array}{rcl}
\denc{\Box t} &=& \nodet{\den{t}}{\phantom{\den{t}}} \\
\denc{\tau_1 + \tau_2} &=& \denc{\tau_1} \oplus \denc{\tau_2} \\
\denc{\tau_1 * \tau_2} &=& \denc{\tau_1} \otimes \denc{\tau_2} \\
\denc{- \tau} &=& \ominus \denc{\tau}
\end{array}\]
where the operations $\oplus$, $\otimes$, and $\ominus$ are defined below.
It is possible, in general, for the binary trees representing the cubes not
be balanced. It is however always possible to embed an $n$-dimensional cube
into a cube of a larger dimension using the fact (to be justified using a
$\Pi$-combinator that witnesses this isomorphism in the next section) that
$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
\[
S = \nodet{S}{\phantom{S}} = 
  \nodet{\nodet{S}{\phantom{S}}}
        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
\]
In the recursive definition of addition below, we assume that the smaller
cube has been extended to the dimension of the larger one: 
\[\begin{array}{rcl}
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} \\
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt}
\end{array}\]

@@

\noindent The definition of negation is natural: it recursively swaps the
positive and negative subspaces. We note that the general definitions of sums
and products reduce to the conventional sums and products for $0$-dimensional
types. Also the sum of 1-dimensional types reduces to the sum used in the
\textbf{Int} construction and that addition and negation do not increase the
dimensions. The interesting new development is the definition of
products. The example in Fig.~\ref{mult} illustrates a general pattern: the
product of an $n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

@@

@@

The terms of $\Pi$ witness type isomorphisms of the form $b \iso b$. They
consist of base isomorphisms, as defined below, and their composition.
\[\begin{array}{rrcll}
\identlp :&  0 + b & \iso & b &: \identrp \\
\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
\identlt :&  1 * b & \iso & b &: \identrt \\
\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
\distz :&~ 0 * b & \iso & 0 &: \factorz \\
\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor
\end{array}\]
Each line of the above table introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of $\Pi$. Note how the above has two readings:
first as a set of typing relations for a set of constants. Second, if these
axioms are seen as universally quantified, orientable statements, they also
induce transformations of the (traditional) values. The (categorical or
homotopical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

\begin{table}[t]
\hrule\medskip
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\quad 
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\quad
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\quad 
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\caption{Combinators\label{pi-combinators}}
\medskip\hrule
\end{table}

>>>>

\newcommand{\bdim}[1]{\textsf{dim}(#1)}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\proves}{\vdash}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}
\newcommand{\adjoint}[1]{#1^{\dagger}}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\plus}{\raisebox{.4\height}{\scalebox{.6}{+}}}
\newcommand{\minus}{\raisebox{.4\height}{\scalebox{.8}{-}}}
\newcommand{\mm}{\texttt{-}}
\newcommand{\pp}{\texttt{+}}
\newcommand{\negp}[1]{\textit{neg}(#1)}
\newcommand{\inl}[1]{\textsf{inl}~#1}
\newcommand{\inr}[1]{\textsf{inr}~#1}
\newcommand{\lolli}{\multimap} 
\newcommand{\cubt}{\mathbb{T}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\nodet}[2]{\fcolorbox{black}{white}{$#1$}\fcolorbox{black}{gray!20}{$#2$}}

@@

%% double-blind reviewing...
\title{Polarized Cubical Types}
\authorinfo{}{}{}
\maketitle

@@

Our cubical types $\tau$ include the empty type 0, the unit type 1,
conventional sum and product types, as well as \emph{negative} types:
\[\begin{array}{rcl}
\tau &::=& 0 \alt 1 \alt \tau_1 + \tau_2 \alt \tau_1 * \tau_2 \alt - \tau
\end{array}\]
We use $\tau_1 - \tau_2$ to abbreviate $\tau_1 + (- \tau_2)$ and more
interestingly $\tau_1 \lolli \tau_2$ to abbreviate $(- \tau_1) + \tau_2$.
The \emph{dimension} of a type is defined as follows:
\[\begin{array}{rcl}
\bdim{0} &=& 0 \\
\bdim{1} &=& 0 \\
\bdim{\tau_1 + \tau_2} &=& \max(\bdim{\tau_1},\bdim{\tau_2}) \\
\bdim{\tau_1 * \tau_2} &=& \bdim{\tau_1} + \bdim{\tau_2} \\
\bdim{- \tau} &=& \max(1,\bdim{\tau})
\end{array}\]
The base types have dimension 0. Using negative types but no product types
raises the dimension to 1. This is the situation with the \textbf{Int} or
$\mathcal{G}$ construction. Once product types are used, the dimension can
increase without bounds.

@@

@@

This point is made precise in the denotation of types which maps a type of
dimension $n$ to $n$-dimensional cube. We represent such a cube syntactically
as a binary tree of maximum depth $n$ with nodes of the form
$\nodet{\cubt_1}{\cubt_2}$. In such a node, $\cubt_1$ is the positive
subspace and $\cubt_2$ (shaded in gray) is the negative subspace along the
first dimension. Each of these subspaces is itself a cube of a lower
dimension. The $0$-dimensional cubes are the conventional denotation of
first-order types as sets. We use $S$ to denote the denotations of such
types. For example, 1-dimensional cube looks like $\nodet{S_1}{S_2}$ which
intuitively corresponds to the difference $t_1 - t_2$ of the two types whose
denotations are $S_1$ and $S_2$ respectively. A full 2-dimensional cube looks
like $\nodet{(\nodet{S_1}{S_2})}{(\nodet{S_3}{S_4})}$ which intuitively
corresponds to the iterated difference of the appropriate types
$(t_1-t_2)-(t_3-t_4)$ where the product of the successive signs from the
outermost box encodes the sign.

Formally, the denotation of types is as follows:
\[\begin{array}{rcl}
\den{0} &=& \emptyset \\
\den{1} &=& \{ \star \} \\
\den{\tau_1 + \tau_2} &=& \den{\tau_1} \oplus \den{\tau_2} \\
\den{\tau_1 * \tau_2} &=& \den{\tau_1} \otimes \den{\tau_2} \\
\den{- \tau} &=& \ominus \den{\tau} \\
\\
\noalign{\mbox{where:}\hfill}
\\
S_1 \oplus S_2 &=& S_1 \uplus S_2 \\
S \oplus (\nodet{\cubt_1}{\cubt_2}) &=& 
  (\nodet{S}{\phantom{S}}) \oplus (\nodet{\cubt_1}{\cubt_2}) \\
(\nodet{\cubt_1}{\cubt_2}) \oplus S &=& 
  (\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{S}{\phantom{S}}) \\
(\nodet{\cubt_1}{\cubt_2}) \oplus (\nodet{\cubt_3}{\cubt_4}) &=& 
  \nodet{\cubt_1 \oplus \cubt_3}{\cubt_2 \oplus \cubt_4} \\
\\
S_1 \otimes S_2 &=& S_1 \times S_2 \\
S \otimes (\nodet{\cubt_1}{\cubt_2}) &=& 
  \nodet{S \otimes \cubt_1}{S \otimes \cubt_2} \\
(\nodet{\cubt_1}{\cubt_2}) \otimes \cubt &=& 
  \nodet{\cubt_1 \otimes \cubt}{\cubt_2 \otimes \cubt} \\
\\
\ominus~S &=& \nodet{\phantom{S}}{S} \\
\ominus~\nodet{\cubt_1}{\cubt_2} &=& \nodet{\ominus~\cubt_2}{\ominus~\cubt_1} 
\end{array}\]
The type 0 maps to the empty set. The type 1 maps to a singleton set. The sum
of $0$-dimensional types is the disjoint union as expected. For cubes of
higher dimensions, the subspaces are recursively added. If the two types to
be added are of different dimensions, it is necessary to embed the smaller
type in the larger dimension using the fact (to be justified using a
$\Pi$-combinator that witnesses this isomorphism in the next section) that
$t$ is equivalent to $t-0$ and to $(t-0)-0$ and so on, i.e.,
\[
S = \nodet{S}{\phantom{S}} = 
  \nodet{\nodet{S}{\phantom{S}}}
        {\phantom{{\nodet{S}{\phantom{S}}}}} = \ldots \\
\]
The product of 0-dimensional types is the cartesian product of sets. The
definition of negation is natural: it recursively swaps the positive and
negative subspaces. We note that the general definitions of sums and products
reduce to the conventional sums and products for $0$-dimensional types. Also
the sum of 1-dimensional types reduces to the sum used in the \textbf{Int}
construction and that addition and negation do not increase the
dimensions. The interesting new development is the definition of
products. The example in Fig.~\ref{mult} illustrates the general pattern: the
product of an $n$-dimensional type with an $m$-dimensional type gives an
$(n+m)$-dimensional type. For low dimensions such as the ones used in the
example above, it is possible to visualize what is happening in coordinate
space (see Fig.~\ref{multv}).

\paragraph*{Note.} 
Our (simple) semantics of types identifies several structurally different
types such as $(1+(1+1))$ and $((1+1)+1)$. In some sense, this is innocent as
the types are isomorphic. However, in the operational semantics discussed in
the next section, we make the computational content of such type isomorphisms
explicit.

@@

The terms of $\Pi$ witness type isomorphisms of the form $b \iso b$. They
consist of base isomorphisms, as defined below, and their composition. Each
line of the above table introduces a pair of dual constants\footnote{where
  $\swapp$ and $\swapt$ are self-dual.} that witness the type isomorphism in
the middle.  These are the base (non-reducible) terms of the second,
principal level of $\Pi$. Note how the above has two readings: first as a set
of typing relations for a set of constants. Second, if these axioms are seen
as universally quantified, orientable statements, they also induce
transformations of the (traditional) values. The (categorical or homotopical)
intuition here is that these axioms have computational content because they
witness isomorphisms rather than merely stating an extensional equality.

@@

\begin{table*}[t]
\[\begin{array}{cc}
\begin{array}{rrcll}
\identlp :&  0 + b & \iso & b &: \identrp \\
\swapp :&  b_1 + b_2 & \iso & b_2 + b_1 &: \swapp \\
\assoclp :&  b_1 + (b_2 + b_3) & \iso & (b_1 + b_2) + b_3 &: \assocrp \\
\identlt :&  1 * b & \iso & b &: \identrt \\
\swapt :&  b_1 * b_2 & \iso & b_2 * b_1 &: \swapt \\
\assoclt :&  b_1 * (b_2 * b_3) & \iso & (b_1 * b_2) * b_3 &: \assocrt \\
\distz :&~ 0 * b & \iso & 0 &: \factorz \\
\dist :&~ (b_1 + b_2) * b_3 & \iso & (b_1 * b_3) + (b_2 * b_3)~ &: \factor \\
\eta :&~ 0 & \iso & b + (-b)~ & : \epsilon
\end{array}
& 
\begin{minipage}{0.5\textwidth}
\begin{center} 
\Rule{}
{}
{\jdg{}{}{\idc : b \iso b}}
{}
\qquad\qquad
\Rule{}
{\jdg{}{}{c : b_1 \iso b_2}}
{\jdg{}{}{\symc{c} : b_2 \iso b_1}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_2 \iso b_3}
{\jdg{}{}{c_1 \fatsemi c_2 : b_1 \iso b_3}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \oplus c_2 : b_1 + b_3 \iso b_2 + b_4}}
{}
\\ \bigskip
\Rule{}
{\jdg{}{}{c_1 : b_1 \iso b_2} \quad c_2 : b_3 \iso b_4}
{\jdg{}{}{c_1 \otimes c_2 : b_1 * b_3 \iso b_2 * b_4}}
{}
\end{center}
\end{minipage}
\end{array}\]
\caption{Combinators\label{pi-combinators}}
\end{table*}
